Coupling finite and boundary element methods to solve the Poisson--Boltzmann equation for electrostatics in molecular solvation

### 2023-05-02 intro.tex
intro and methods section

In biologically relevant settings, the structure and function of biomolecules are largely determined by the surrounding water, which usually contains salt. 
To describe these systems accurately, we need to account for the solvent correctly, which has given rise to a wide range of models.\cite{onufriev2018water}
Highly detailed models consider every water molecule and salt ion explicitly, however, there are approximated models that use continuum theory to represent this ionic solution, known as implicit-solvent models.\cite{RouxSimonson1999,DecherchiETal2015}
In the case of electrostatics, the implicit-solvent model is mathematically characterized by the Poisson-Boltzmann equation (PBE)\cite{Baker2004,Bardhan2012}, which is widely used to compute solvation free energies and mean-field potentials.

The implicit-solvent model for electrostatics describes the dissolved molecule as an infinite medium with a low-dielectric solute-shaped cavity, which contains a charge distribution from the partial charges --- usually a sum of Dirac deltas at the atom's locations.
The outer solvent region is represented with a high dielectric, and considers the presence of salt.
These two regions are interfaced by the molecular surface, which can be defined in various ways \cite{HarrisBoschitcshFenley2013}, where the continuity of the electrostatic potential and electric displacement are enforced.

The PBE has been solved numerically with finite difference\cite{BakerETal2001,GilsonETal1985,JurrusETal2018,LiETal2019}, finite element\cite{HolstETal2012,BondEtal2010}, boundary element\cite{boschitsch2002fast,LuETal2006,AltmanBardhanWhiteTidor09,bajaj2011efficient,GengKrasny2013,CooperBardhanBarba2014}, and analytical\cite{YapHeadgordon2010,FelbergETal2017} methods.
In particular, the boundary element method (BEM) has proven to be very efficient for high-accuracy calculations \cite{GengKrasny2013,CooperBardhanBarba2014}, mainly due to the precise description of the molecular surface and point charges. 
However, BEM is limited to constant material properties in each region and the linear version of the PBE. 
Even though these limitations are acceptable in a wide range of applications, there are cases when BEM falls short, for example, if a variable permittivity is required inside the solute \cite{grant2001smooth,li2013dielectric}, or the solute is highly charged such that the linear approximation breaks\cite{FogolariETal1999}.

The present article describes a methodology to overcome some of those limitations, by coupling finite and boundary elements methods.
This approach brings the best of both worlds: the flexibility of FEM and the efficiency of BEM, all in an accurate description of the solute molecule.
FEM-BEM coupling is a popular technique in the context of mixed linear-nonlinear models,\cite{carstensen1995coupling,aurada2013classical} fracture mechanics,\cite{aour2007coupled} fluid-structure interaction,\cite{estorff1991fem} acoustics,\cite{hiptmair2006stabilized} and electromagnetics.\cite{matsuoka1988calculation,hiptmair2008stabilized,bruckner20123d}
On the other hand, the PBE has been solved with hybrid numerical methods in the past, for example, by coupling finite difference with boundary elements\cite{boschitsch2004hybrid} or finite elements,\cite{xie2016new,ying2018hybrid} which were mainly used to solve the nonlinear PBE in a specific region, and to implement modifications to the PBE model ({\it ie.} the size-modified PBE).
To the best of our knowledge, this is the first time finite and boundary elements are combined in this application.

Here we prototype this principle with the simplest implementation possible: a Johnson-N\'ed\'elec\cite{johnson1980coupling} coupling, mass-matrix and diagonal preconditioning, and no distributed memory execution. 
This limits the size of problems we can access currently, however, it sets the basis for future developments that use more elaborate formulations and algorithms that are readily available in the software libraries used in this work, in particular, Bempp-cl and FEniCSx.
These libraries are designed for ease of use through a simple Python interface, making them ideal tools to explore and improve computational efficiency, towards tackling large-scale problems, such as a full viral capsid.\cite{MartinezETal2019,wang2021high}
We hope this work will inspire research along these lines.

>>>>

In biologically relevant settings, the structure and function of biomolecules are largely determined by the surrounding water, which usually contains salt. 
To describe these systems accurately, we need to account for the solvent correctly, which has given rise to a wide range of models.\cite{onufriev2018water}
Highly detailed models consider every water molecule and salt ion explicitly.
For solutions with large numbers of molecules, however, these models can be very compuationally expensive, so implicit-solvent models---approximated models that use continuum theory to represent the ionic solution---are often used instead.\cite{RouxSimonson1999,DecherchiETal2015}
In the case of electrostatics, the implicit-solvent model is mathematically characterized by the Poisson--Boltzmann equation (PBE)\cite{Baker2004,Bardhan2012}, which is widely used to compute solvation free energies and mean-field potentials.

The implicit-solvent model for electrostatics describes the dissolved molecule as an infinite medium with a low-dielectric solute-shaped cavity, which contains a charge distribution from the partial charges---usually a sum of Dirac deltas at the atom's locations.
The outer solvent region is represented with a high dielectric constant, and considers the presence of salt.
These two regions are interfaced by the molecular surface where the continuity of the electrostatic potential and electric displacement are enforced.
The molecular surface can be defined in various ways \cite{HarrisBoschitcshFenley2013}.

The PBE has been solved numerically with finite difference\cite{BakerETal2001,GilsonETal1985,JurrusETal2018,LiETal2019}, finite element\cite{HolstETal2012,BondEtal2010}, boundary element\cite{boschitsch2002fast,LuETal2006,AltmanBardhanWhiteTidor09,bajaj2011efficient,GengKrasny2013,CooperBardhanBarba2014}, and analytical\cite{YapHeadgordon2010,FelbergETal2017} methods.
In particular, the boundary element method (BEM) has proven to be very efficient for high-accuracy calculations \cite{GengKrasny2013,CooperBardhanBarba2014}, mainly due to the precise description of the molecular surface and point charges. 
However, BEM is limited to constant material properties in each region and the linear version of the PBE. 
Even though these limitations are acceptable in a wide range of applications, there are cases when BEM falls short: for example, if a variable permittivity is required inside the solute \cite{grant2001smooth,li2013dielectric}, or the solute is highly charged such that the linear approximation breaks\cite{FogolariETal1999}.

The present article describes a methodology to overcome some of those limitations, by coupling finite and boundary element methods.
This approach brings the best of both worlds---the flexibility of FEM and the efficiency of BEM---all in an accurate description of the solute molecule.
FEM-BEM coupling is a popular technique in the context of mixed linear-nonlinear models,\cite{carstensen1995coupling,aurada2013classical} fracture mechanics,\cite{aour2007coupled} fluid-structure interaction,\cite{estorff1991fem} acoustics,\cite{hiptmair2006stabilized} and electromagnetics.\cite{matsuoka1988calculation,hiptmair2008stabilized,bruckner20123d}
On the other hand, the PBE has been solved with hybrid numerical methods in the past: for example, by coupling finite differences with boundary elements\cite{boschitsch2004hybrid} or finite elements\cite{xie2016new,ying2018hybrid} to solve the nonlinear PBE in a specific region, and to implement modifications to the PBE model ({\it ie.} the size-modified PBE).
To the best of our knowledge, this is the first time finite and boundary elements have been combined in this application.

In this paper, we prototype this principle with the simplest implementation possible: a Johnson--N\'ed\'elec\cite{johnson1980coupling} coupling, where we solve using mass-matrix and diagonal preconditioning and whithout distributed memory execution.
This limits the size of problems we can access currently, however, it sets the basis for future developments that use more elaborate formulations and algorithms that are readily available in open source software libraries.
In this work, we use the boundary element library Bempp-cl\cite{BetckeScroggs2021} and the finite element library FEniCSx\cite{BasixJoss,BasixDofTransformations,FEniCS}.
These libraries are easy to use and their full functionalities may be accessed via their Python interfaces, making them ideal tools for easily implementing problems like this as well as for exploring computational efficiency and moving towards tackling large-scale problems, such as a full viral capsid.\cite{MartinezETal2019,wang2021high}
We hope this work will inspire research along these lines.

### 2023-05-02 main.tex
Updated results, comparison for tol = 1e-4 and 1e-5
intro and methods section

@@

\title{Coupling finite and boundary element methods to solve the Poisson-Boltzmann equation for electrostatics in molecular solvation}
\author{Micha\l{} Bosy\thanks{School of Computer Science and Mathematics, Kingston University London, Penrhyn Road, Kingston upon Thames, KT1 2EE, UK}, 
    Matthew W. Scroggs\thanks{Department of Mathematics, University College London, 25 Gordon Street, WC1H 0AY London, UK}, 
    Timo Betcke\footnotemark[2],~\\%\thanks{Department of Mathematics, University College London, 25 Gordon Street, WC1H 0AY London, UK}, 
    Erik Burman\footnotemark[2],~%\thanks{Department of Mathematics, University College London, 25 Gordon Street, WC1H 0AY London, UK}, 
    Christopher D. Cooper\thanks{Department of Mechanical Engineering and Centro Cient\'ifico Tecnol\'ogico de Valpara\'iso, Universidad T\'ecnica Federico Santa Mar\'ia, Valpara\'iso 2390123, Chile}}

@@

\begin{abstract}
The Poisson-Boltzmann equation is widely used to model electrostatics in molecular systems. Available software packages solve it using finite difference, finite element, and boundary element methods, where the latter is attractive due to the accurate representation of the molecular surface and partial charges, and exact enforcement of the boundary conditions at infinity. However, the boundary element method is limited to linear equations and piece-wise constant variations of the material properties. In this work, we present a scheme that couples finite and boundary elements for the Poisson-Boltzmann equation, where the finite element method is applied in a confined {\it solute} region, and the boundary element method in the external {\it solvent} region. As a proof-of-concept exercise, we used the simplest methods available: Johnson-N\'ed\'elec coupling with mass matrix and diagonal preconditioning, implemented in the Bempp-cl and FEniCSx libraries from a simple Python interface. We showcased our implementation by computing the polar component of the solvation free energy of a set of molecules using a constant and a Gaussian-varying permittivity. First, we validated our implementation against the finite difference code APBS (to 0.5\%), and then, we show scaling from protein G B1 (955 atoms) up to immunoglobulin G (20 148 atoms). For small problems we saw that the coupled method was efficient, outperforming a purely boundary integral approach, but scaled worse with size. For Gaussian-varying permittivities, which are beyond the applicability of boundary elements, we were able to run medium to large problems on a single workstation. However, our results illustrate the need to improve conditioning and incorporate distributed memory parallelism as we aim at larger systems. We hope this work will serve as inspiration for future developments that consider space-varying field parameters, and mixed linear-nonlinear schemes for molecular electrostatics with implicit solvent models. 
\end{abstract}

\begin{wileykeywords}
Finite element method, Boundary element method, Poisson-Boltzmann, Implicit solvent model, Electrostatics.
\end{wileykeywords}

@@

\begin{figure}[h]
\centering
\colorbox{background-color}{
\fbox{
\begin{minipage}{1.0\textwidth}
\includegraphics[width=50mm,height=50mm]{graphical_abstract_fem-bem.png} %Pick only one of the two styles by uncommenting the corresponding \includegraphics
%\includegraphics[width=110mm,height=20mm]{cc.eps}
\\
    The boundary element method is a popular numerical algorithm to solve the Poisson-Boltzmann equation for molecular electrostatics. However, this technique is limited to linear equations and piece-wise constant variations of the field parameters. Here, we overcome such limitations by coupling it with a finite element method, implemented using the Bempp-cl and FEniCSx libraries through simple Python interfaces. This results in an efficient, accurate, flexible, and easy-to-use computational tool for model development. 
\end{minipage}
}}
\end{figure}

% makes references listed with 1., 2., etc.  
  \makeatletter
  \renewcommand\@biblabel[1]{#1.}
  \makeatother

@@

\subsection*{\sffamily \large ACKNOWLEDGMENTS}

CDC acknowledges the support from CCTVal through ANID PIA/APOYO AFB220004.
%((Place Acknowledgments here))

>>>>

\newcommand{\mscomment}[1]{\textbf{\color{red}[Matthew: #1]}}

@@

\title{Coupling finite and boundary element methods to solve the Poisson--Boltzmann equation for electrostatics in molecular solvation}
\author{Micha\l{} Bosy\thanks{School of Computer Science and Mathematics, Kingston University London, Penrhyn Road, Kingston upon Thames, KT1 2EE, UK}, 
    Matthew W. Scroggs\thanks{Department of Mathematics, University College London, 25 Gordon Street, WC1H 0AY London, UK}, 
    Timo Betcke\footnotemark[2],~\\%\thanks{Department of Mathematics, University College London, 25 Gordon Street, WC1H 0AY London, UK}, 
    Erik Burman\footnotemark[2],~%\thanks{Department of Mathematics, University College London, 25 Gordon Street, WC1H 0AY London, UK}, 
    Christopher D. Cooper\thanks{Department of Mechanical Engineering and Centro Cient\'ifico Tecnol\'ogico de Valpara\'iso, Universidad T\'ecnica Federico Santa Mar\'ia, Valpara\'iso 2390123, Chile}}

@@

\begin{abstract}
The Poisson--Boltzmann equation is widely used to model electrostatics in molecular systems. Available software packages solve it using finite difference, finite element, and boundary element methods, where the latter is attractive due to the accurate representation of the molecular surface and partial charges, and exact enforcement of the boundary conditions at infinity. However, the boundary element method is limited to linear equations and piecewise constant variations of the material properties. In this work, we present a scheme that couples finite and boundary elements for the Poisson--Boltzmann equation, where the finite element method is applied in a confined {\it solute} region, and the boundary element method in the external {\it solvent} region. As a proof-of-concept exercise, we use the simplest methods available: Johnson--N\'ed\'elec coupling with mass matrix and diagonal preconditioning, implemented using the Bempp-cl and FEniCSx libraries via their Python interfaces. We showcase our implementation by computing the polar component of the solvation free energy of a set of molecules using a constant and a Gaussian-varying permittivity. We validate our implementation against the finite difference code APBS (to 0.5\%), and show scaling from protein G B1 (955 atoms) up to immunoglobulin G (20\,148 atoms). For small problems, the coupled method was efficient, outperforming a purely boundary integral approach. For Gaussian-varying permittivities, which are beyond the applicability of boundary elements alone, we were able to run medium to large sized problems on a single workstation. Development of better preconditioning techniques and the use of distributed memory parallelism for larger systems remains an area for future work. We hope this work will serve as inspiration for future developments that consider space-varying field parameters, and mixed linear-nonlinear schemes for molecular electrostatics with implicit solvent models. 
\end{abstract}

\begin{wileykeywords}
Finite element method, Boundary element method, Poisson--Boltzmann, Implicit solvent model, Electrostatics.
\end{wileykeywords}

@@

\begin{figure}[h]
\centering
\colorbox{background-color}{
\fbox{
\begin{minipage}{1.0\textwidth}
\includegraphics[width=50mm,height=50mm]{graphical_abstract_fem-bem.png} %Pick only one of the two styles by uncommenting the corresponding \includegraphics
%\includegraphics[width=110mm,height=20mm]{cc.eps}
\\
    The boundary element method is a popular numerical algorithm to solve the Poisson--Boltzmann equation for molecular electrostatics. However, this technique is limited to linear equations and piecewise constant variations of the field parameters. Here, we overcome such limitations by coupling it with a finite element method, implemented using the Bempp-cl and FEniCSx libraries via their Python interfaces. This results in an efficient, accurate, flexible, and easy-to-use computational tool for model development. 
\end{minipage}
}}
\end{figure}

% makes references listed with 1., 2., etc.  
0  \makeatletter
  \renewcommand\@biblabel[1]{#1.}
  \makeatother

@@

\subsection*{\sffamily \large ACKNOWLEDGMENTS}
MB acknowledges the support from Kingston University through First Kingston University Grant. \\
CDC acknowledges the support from CCTVal through ANID PIA/APOYO AFB220004.
%((Place Acknowledgments here))
